
\section{Plan} \label{plan}

I will now lay out my plan for the remainder of my Ph.D. research. 
% 
I will give a high level view, with some details as to how I intend to address my research problems. 
% 
I will also provide a gantt chart, showing my progress so far and how my future work fits in to this.


\paragraph{Year 2}
For the remainder of my second year, I  have two main goals.
% 
Firstly, I aim to resubmit my survey paper, either at a new venue, or again at Transactions in Quantum Computing.
% 
This requires some substantial revisions as indicated by the reviewers, including the addition of several languages to the survey,
as well as reworking the example programs used throughout.
% 
Therefore, I aim to have this done by October.


Secondly, I aim to continue my work on the mechanising hybrid language semantics.
% 
I believe the most difficult part of this will be figuring out a suitable representation
of the state at higher levels in the CCF lattice.
% 
I aim to complete at least the semantics for the next language in the lattice (Cond) by October.
% 
I will aim to have a workshop paper on the topic submitted before the end of the year.

\paragraph{Year 3}
In my third year and beyond, I will aim to employ the mechanised semantics to build
verified optimisation passes.
% 
This will involve investigating the ``optimisation lifting'' technique we have already discussed, as well as 
investigating the specification and verification of some simple rewrite rules. 



\subsection{Timeline \& Gantt Chart}

\input{report/diagrams/gantt-chart}

A timeline of this plan of work is included in \cref{chart}.
