
\section{Other Activities} \label{outreach}

In the course of the past 9 months I have also taken part in various
other activities, which I have listed below:

\paragraph{SPLS}
I attend the Scottish Programming Languages Seminar (SPLS), whenever it is held.
So far this year I have attended seminars in St Andrews and Edinburgh.

\paragraph{POPL `26}
In January I attended The 53rd ACM SIGPLAN Symposium on Principles of Programming Languages (POPL 2026) in Rennes, France. 
% 
While there, I also attended the Programming Languages Mentoring Workshop (PLMW).
% 
I found this to be a wonderful way to understand what state-of-the art programming language research looks like,
and to network with my peers.


\paragraph{Systems coffee \& reading group}
In January, I was asked to take over the running of the Glasgow systems section coffee morning.
Thus runs every week on wednesdays at 10:30 AM.

\paragraph{Tutoring}
I am involved in undergraduate teaching by way of my role as a tutor/GTA.
This year I taught Functional programming in the first semester. In the second semester I taught Quantum Computing

\paragraph{PLUG seminars \& PLunch}
Glasgow's programming language theme holds a series of seminars.
% 
I make an effort to attend these whenever possible. 
% 
I gave a talk there in the last year, titled ``Package Managers, through a lens or two''. 
% 
I hope to continue the work presented there at some point, though it is not directly relevant to my PhD at this time.


The group also takes lunch together approximately once a month as a social event. 
% 
I also try to attend these.

\paragraph{SPLV Organisation Committee}
This year I have served on the Scottish Programming Languages and Verification Summer School's organising committee,
as it will be held in Glasgow later in August.
% 
I have been mostly in charge of tracking our accounts.

\paragraph{QI CDT Spring School}
In May, I attended the Quantum Informatics CDT Spring School on the Isle of Skye.
% 
I found this to be extremely helpful in connecting me with my peers working in quantum computation more broadly.
% 
I also found the lectures there to be very helpful reinforcing some of the basics I had already covered earlier in my PhD.

\paragraph{FATA Social Events}
I have been involved in several events organised by the formal analysis, theory and algorithms section here at Glasgow.
These include the away day, the new running group, and the cakes/coffee afternoon on Mondays.
